# Preface {.unnumbered}

My notes on learning `data.table`.

# Introduction to data.table

>
if you are interested in reducing **programming** and **compute** time tremendously, then this package is for you
>

NB: Two type of **efficiency**: coder's efficiency and code's efficiency --- reference: 
Colin Gillespie, C.; Lovelace, R. (2021). Efficient R programming. URL: https://csgillespie.github.io/efficientR/


File reader `fread`

- data.table's fast-and-friendly file reader `fread` 
- `fread` accepts http and https URLs directly

```{r, warning=FALSE, message=FALSE, eval=FALSE}
library(data.table)
link <- "https://raw.githubusercontent.com/Rdatatable/data.table/master/vignettes/flights14.csv"
flights <- fread(link)
```

## Basics

- `data.table()`

```{r}
library(data.table)
small_dtb <- data.table(x = letters[1:5],
                        y = rnorm(5),
                        z = 1:5)
small_dtb
```

>
You can also convert existing objects to a data.table using `setDT()` (for data.frame and list structures) or `as.data.table()` (for other structures).
>

- general form `DT[i, j, by]` --- **Take DT, subset/reorder rows using i, then calculate j, grouped by by.**

- subset rows in i

**Example:**
```{r}
(small_dtb[x %in% c('a', 'b', 'c', 'e') & z >= 3])
```

**Example:**
```{r}
(small_dtb[c(2, 5)])
```

up to – Sort flights first by column origin in ascending order, and then by dest in descending order: