# Secondary indices and auto indexing

https://cran.r-project.org/web/packages/data.table/vignettes/datatable-secondary-indices-and-auto-indexing.html

## What 

>
Secondary indices are similar to keys in data.table, except for two major differences:

>
- It doesn't physically reorder the entire data.table in RAM. Instead, it only computes the order for the set of columns provided and stores that order vector in an additional attribute called index.

>
- There can be more than one secondary index for a data.table (as we will see below).
>

## How

- `setindex()` and `setindexv()` allows adding a secondary index to the data.table.

- The function `indices()` returns all current secondary indices in the data.table. If none exists, `NULL` is returned.

- `setindex(dt, NULL)` would remove all secondary indices

- **on argument**
    - enables subsetting by computing secondary indices on the fly. This eliminates having to do `setindex()` every time.
    - allows easy reuse of existing indices by just checking the attributes.
    - allows for a cleaner syntax by having the columns on which the subset is performed as part of the syntax. This makes the code easier to follow when looking at it at a later point.

## Examples

```{r, warning=FALSE, message=FALSE}
library(data.table)

set.seed(123)
dt <- data.table(ID = 1:1000,
                 month = sample(month.name, 1000, replace = TRUE),
                 x = sample(letters, 1000, replace = TRUE),
                 y = runif(1000))

# Indexed subsetting (explicit column specification)
setindex(dt, x)
dt[.(letters[1:3]), on = 'x'][]

(names(attributes(dt)))

(indices(dt))

```

**Use `on` argument**

```{r}
setindex(dt, NULL)
(indices(dt))

dt[.(c('a', 'b'), month.name[c(1, 12)]), on = c('x', 'month'), 
   nomatch = NULL][]
# or for clarity/readability
dt[.(x = c('a', 'b'), month = month.name[c(1, 12)]), on = c('x', 'month'), 
   nomatch = NULL][]

# comparing 
dt[x %in% c('a', 'b') & month %in% month.name[c(1, 12)]][]

```

```{r}
# Compute or do in j
dt[.(x = c('a', 'b'), month = month.name[c(1, 12)]), 
   mean(y),
   on = c('x', 'month'),]
```

```{r}
# sub-assign by reference using := in j
dt[.(month = "December"), month := "Dec", on = 'month']
dt[, (nbr_obs = .N), by = month][]
```


```{r}
# change back from 'Dec' to "December"
dt[.(month = "Dec"), month := "December", on = 'month']
# Aggregation using by
dt[.(x = 'a'), .(mean_of_y = mean(y)), keyby = month, on = 'x'][
  order(match(month, month.name))][]
```

```{r}
# The mult argument
dt[.(x = c('a', 'b', 'c')), on = 'x', mult = 'first'][]
dt[.(x = c('a', 'b', 'c')), on = 'x', mult = 'last'][]

```

```{r}
# The nomatch argument
dt[.(x = 'a', month = c('Dec')), on = c("x", "month")][]

dt[.(x = 'a', month = c('Dec')), on = c("x", "month"), nomatch = NULL][]

```

